### 2022-04-01 Acta Informatica2022/5_complexity.tex
before modify sec 4

\begin{lemma}\label{lem:cpx exp}
    \begin{align}
        c'_0&\le c_0^2 \notag \\
        s'_0&\le ms_0^2 \notag\\
        a'_0&\le s_0a_0\notag 
    \end{align} 
\end{lemma}

\begin{proof}
The analysis is divided into two cases by assuming all atomic formulas are of the same form (inequalities atoms or divisibility atoms).

If all atoms are inequalities of the form in Lemma~\ref{lem:exp-ineq}. We know that each atomic formula $\tau$ with exponential occurrence of $\ivarx_m$ is replaced by a new formula. Note that coefficients of linear occurrences of $\ivarx_i$, for $i\le m-2$, remain changed throughout the substitutions, only the coefficients of linear occurrences of $\ivarx_m$ and $\ivarx_{m-1}$ will be changed: constant 1 is introduced as a coefficient for $\ivarx_m$, and if we substitute $\ivarx_m$ by $\ivarx_{m-1}+p$ for some constant $p$, coefficient of linear occurrence of $\ivarx_{m-1}$ will become $b_m+b_{m-1}$ ($b_m,b_{m-1}$ are coefficients for $\ivarx_{m}$ and $\ivarx_{m-1}$ in $\tau$, see Lemma~\ref{lem:exp-ineq} ). Since the new coefficient is obtained by adding two linear coefficient together, we have $c'_1\le c_0^2$. Note that $\delta$ and $B$ in Lemma~\ref{lem:exp-ineq} are at most $\ell_{10}(ms_0)$. When we substitute $\ivarx_m$ by $\ivarx_{m-1}+\delta-1$ or by $B-1$, the largest constant in the formula becomes at most $s_0\cdot  10^{\ell_{10}(ms_0)}\le ms_0^2$. And an inequality is replaced by at most $4\ell_{10}(ms_0)$ atomic formulas, so $a'_1\le 4\ell_{10}(ms_0)a_0$.

@@

Choose larger upper bounds for $c'_1$, $s'_1$ and $a'_1$ respectively, then the lemma is proved.

\end{proof}

@@

\begin{lemma}\cite{Oppen73}\label{lem:cpx oppen}
    \begin{align}
        c_1&\le c_0'^4 \notag \\
        s_1&\le s_0'^{4c_0'} \notag\\
        a_1&\le {a_0'}^4 s_0'^{2c_0'}\notag 
    \end{align} 
\end{lemma}

Combining Lemma~\ref{lem:cpx exp} and Lemma~\ref{lem:cpx oppen}, we have
\begin{align}
    c_1&\le c_0^8 \notag \\
    s_1&\le (ms_0^2)^{4c_0^2} \notag\\
    a_1&\le (s_0a_0)^4(m_0s_0^2)^{2c_0^2}\notag 
\end{align}

@@

\begin{lemma}
    \begin{align}
        c_k&\le c_0^{8^k} \notag \\
        s_k&\le N^{(4c_0)^{ 8^k}} s_0^{(8c_0)^{ 8^k}} \notag\\
        a_k&\le a_0^{4^k}n^{(4c_0)^{ 8^k}} s_0^{(8c_0)^{ 8^k}} \notag
    \end{align} 
\end{lemma}

If we adopt the assumptions in \cite{Oppen73}, by assuming $c_0\le N$, $a_0\le N$ and $s_0\le N$, the space required to store the quantifier free formula $\varphi_k$, is bounded by the product of the number of linear orders $m!$, the number of atoms $a_k$, the maximum number of constants $2m+2$ per atom, the maximum amount of space $s_k$ to store each constant and some constant $q$. So the space complexity is $O(m! \cdot a_k \cdot (2m+2) \cdot s_k)=O(2^{2^{2^{pn \log n}}})$, which belongs to 3-EXPSPACE.

@@

The above analysis is focused on the existential {\paexp} formulas and is sufficient for our string constraints solving context. But what about general {\paexp} formulas that contain alternating quantifiers? Unfortunately, directly extending our analysis on alternating quantifers by transforming $\forall \ivarx.\varphi$ into $\neg \exists \ivarx.\neg \varphi$ will not give an elementary upper bound. We will explain the main idea without going into details on this issue. To eliminate quantifiers in a formula, say $\forall \ivarx_2\exists \ivarx_1.\varphi(\ivarx_1,\ivarx_2,\vec \ivary)$, we first eliminate $\exists \ivarx_1$, treating $\ivarx_2$ and $\vec \ivary$ both as free variables. The subprocedure to eliminate exponential occurrences of $\ivarx_1$ is at the cost of introducing logarithmic expressions of $\ivarx_2$ and $\vec \ivary$, which in the worst case increases \emph{exponentially} the number of fresh variables in the normalization step w.r.t $\ivarx_2$. We conjecture that the space complexity of quantifier elimination procedure over {\paexp} formulas still has an elementary upper bound, which may require a more detailed analysis or improvements of the algorithm.

>>>>

\begin{lemma}\label{lem:cpx exp}
$$c'_0\le {c_0}^2 \qquad s'_0\le m{s_0}^2 \qquad a'_0\le s_0a_0$$ 
\end{lemma}

\hide{\begin{align}
    c'_0&\le c_0^2 \notag \\
    s'_0&\le ms_0^2 \notag\\
    a'_0&\le s_0a_0\notag 
\end{align}}
\begin{proof}
The analysis is divided into two cases by assuming all atomic formulas are of the same form (inequalities atoms or divisibility atoms).

If all atoms are inequalities of the form in Lemma~\ref{lem:exp-ineq}. We know that each atomic formula $\tau$ with exponential occurrence of $\ivarx_m$ is replaced by a new formula. Note that coefficients of linear occurrences of $\ivarx_i$, for $i\le m-2$, remain unchanged throughout the substitutions, only the coefficients of linear occurrences of $\ivarx_m$ and $\ivarx_{m-1}$ will be changed: constant 1 is introduced as a coefficient for $\ivarx_m$, and if we substitute $\ivarx_m$ by $\ivarx_{m-1}+p$ for some constant $p$, coefficient of linear occurrence of $\ivarx_{m-1}$ will become $b_m+b_{m-1}$ ($b_m,b_{m-1}$ are coefficients for $\ivarx_{m}$ and $\ivarx_{m-1}$ in $\tau$, see Lemma~\ref{lem:exp-ineq} ). Since the new coefficient is obtained by adding two linear coefficient together, we have $c'_1\le {c_0}^2$. Note that $\delta$ and $B$ in Lemma~\ref{lem:exp-ineq} are at most $\ell_{10}(ms_0)$. When we substitute $\ivarx_m$ by $\ivarx_{m-1}+\delta-1$ or by $B-1$, the largest constant in the formula becomes at most $s_0\cdot  10^{\ell_{10}(ms_0)}\le m{s_0}^2$. And an inequality is replaced by at most $4\ell_{10}(ms_0)$ atomic formulas, so $a'_1\le 4\ell_{10}(ms_0)a_0$.

@@

Choose larger upper bounds for $c'_1$, $s'_1$ and $a'_1$ respectively, then the lemma is proved.\end{proof}

@@

\begin{lemma}\cite{Oppen73}\label{lem:cpx oppen}
$$c_1\le {c_0'}^4\qquad
s_1\le {s_0'}^{4c_0'}\qquad
a_1\le {a_0'}^4 {s_0'}^{2c_0'}
$$
\end{lemma}

\hide{    \begin{align}
    c_1&\le {c_0'}^4 \notag \\
    s_1&\le {s_0'}^{4c_0'} \notag\\
    a_1&\le {a_0'}^4 {s_0'}^{2c_0'}\notag 
\end{align} }

Combining Lemma~\ref{lem:cpx exp} and Lemma~\ref{lem:cpx oppen}, we have
$$c_1\le {c_0}^8\qquad
s_1\le (m{s_0}^2)^{4{c_0}^2} \qquad
a_1\le (s_0a_0)^4(m_0{s_0}^2)^{2{c_0}^2} 
$$

\hide{\begin{align}
    c_1&\le {c_0}^8 \notag \\
    s_1&\le (m{s_0}^2)^{4{c_0}^2} \notag\\
    a_1&\le (s_0a_0)^4(m_0{s_0}^2)^{2{c_0}^2}
\end{align} }

@@

\begin{lemma}
$$c_k\le {c_0}^{8^k}\qquad
s_k\le N^{(4c_0)^{ 8^k}} {s_0}^{(8c_0)^{ 8^k}} \qquad
a_k\le {a_0}^{4^k}n^{(4c_0)^{ 8^k}} s_0^{(8c_0)^{ 8^k}} $$
\end{lemma}

\hide{    \begin{align}
    c_k&\le {c_0}^{8^k} \notag \\
    s_k&\le N^{(4c_0)^{ 8^k}} {s_0}^{(8c_0)^{ 8^k}} \notag\\
    a_k&\le {a_0}^{4^k}n^{(4c_0)^{ 8^k}} s_0^{(8c_0)^{ 8^k}} \notag
\end{align} 
}
If we adopt the assumptions in \cite{Oppen73}, by assuming $c_0\le N$, $a_0\le N$ and $s_0\le N$, the space required to store the quantifier free formula $\varphi_k$, is bounded by the product of the number of linear orders $m!$, the number of atoms $a_k$, the maximum number of constants $2m+2$ per atom, the maximum amount of space $s_k$ to store each constant and some constant $q$. So the space complexity is $O(m! \cdot a_k \cdot (2m+2) \cdot s_k)=O(2^{2^{2^{pn \log n}}})$, which belongs to 3-EXPSPACE.

@@

The above analysis is focused on the existential {\paexp} formulas and is sufficient for our string constraints solving context. 
Unfortunately, directly extending our analysis on general {\paexp} formulas with alternating quantifiers by transforming $\forall \ivarx.\varphi$ into $\neg \exists \ivarx.\neg \varphi$ will not give an elementary upper bound.

\hide{We will explain the main idea without going into details on this issue. To eliminate quantifiers in a formula, say $\forall \ivarx_2\exists \ivarx_1.\varphi(\ivarx_1,\ivarx_2,\vec \ivary)$, we first eliminate $\exists \ivarx_1$, treating $\ivarx_2$ and $\vec \ivary$ both as free variables. The subprocedure to eliminate exponential occurrences of $\ivarx_1$ is at the cost of introducing logarithmic expressions of $\ivarx_2$ and $\vec \ivary$, which in the worst case increases \emph{exponentially} the number of fresh variables in the normalization step w.r.t $\ivarx_2$. We conjecture that the space complexity of quantifier elimination procedure over {\paexp} formulas still has an elementary upper bound, which may require a more detailed analysis or improvements of the algorithm.}

### 2022-04-01 Acta Informatica2022/6_optimization.tex
before modify sec 4

Let $\alpha_1 = \ell_{10}(c_1) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c_2) -  \ell_{10}(a_{i-1})$. Then from Lemma~\ref{lem:exp-ineq}, let $A\Def 1+\sum_{j=1}^{i-2}\vert a_j\vert$, 
$B \Def 1+\sum_{j=1}^{i-1}\vert b_j\vert$, 
$\delta\Def  \ell_{10}(A)+2$,
and $\gamma \Def 2\ell_{10}(B)+3$. 
\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \ge \gamma$, $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-lb}), thus also $\tau'$, is evaluated to $\ltrue$,
\item if $\ivarx_{\sigma(i-1)} \ge \gamma$, $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$, then inequality (\ref{eqn-ub}), thus also $\tau'$, is evaluated to $\lfalse$.
\end{itemize}

Therefore, $\tau'$ and $\tau$ are equivalent to 
\begin{equation}\label{lemma-example}\begin{aligned}
    &\bigvee_{p=0}^{\gamma-1} (\ivarx_{\sigma(i-1)}=p \wedge \tau[p/\ivarx_{\sigma(i-1)}])\\
    \vee &\bigvee_{p=0}^{\delta-1} \bigg(\ivarx_{\sigma(i-1)}=\ivarx_{\sigma(i-2)}+p\wedge \tau[\ivarx_{\sigma(i-2)}+p/\ivarx_{\sigma(i-1)}]\wedge \ivarx_{\sigma(i-1)}\le \alpha_1-1\bigg)\\
    \vee &\bigvee_{p=0}^{\delta-1} \bigg(\ivarx_{\sigma(i-1)}=\ivarx_{\sigma(i-2)}+p\wedge \tau[\ivarx_{\sigma(i-2)}+p/\ivarx_{\sigma(i-1)}]\wedge \ivarx_{\sigma(i-1)}\ge \alpha_2+2\bigg)\\
    \vee & \bigg( \ivarx_{\sigma(i-1)}\ge \gamma \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)}+\delta \wedge \ivarx_{\sigma(i-1)}\le \alpha(\vec\ivary)-1\bigg)\\
    \vee & \bigvee_{p=\alpha_1}^{\alpha_2+1} \bigg( \ivarx_{\sigma(i-1)}\ge \gamma \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)}+\delta \wedge \ivarx_{\sigma(i-1)} = p \wedge \tau[p/\ivarx_{\sigma(i-1)}]\bigg),
\end{aligned}\end{equation}
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated.

>>>>

Let $\alpha_1 = \ell_{10}(c_1) -  \ell_{10}(a_{i-1})$ and $\alpha_2 = \ell_{10}(c_2) -  \ell_{10}(a_{i-1})$. 
Then from Lemma~\ref{lem:exp-ineq}, let $A\Def 1+\sum_{j=1}^{i-2}\vert a_j\vert$, 
$B \Def 1+\sum_{j=1}^{i-1}\vert b_j\vert$, 
$\delta\Def  \ell_{10}(A)+2$,
and $\gamma \Def 2\ell_{10}(B)+3$. 
When $\ivarx_{\sigma(i-1)} \ge \gamma$ and $\ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)} + \delta$

\begin{itemize}
\item if $\ivarx_{\sigma(i-1)} \le \alpha_1 - 1$, then inequality (\ref{eqn-lb}), thus also $\tau'$, holds.
\item if $\ivarx_{\sigma(i-1)} \ge \alpha_2 + 2$, then inequality (\ref{eqn-ub}), thus also $\tau'$, \textbf{does not} holds.
\end{itemize}

Therefore, $\tau'$ and $\tau$ are equivalent to 
\begin{equation}\label{opt-example}\begin{aligned}
    &\bigvee_{p=0}^{\gamma-1} (\ivarx_{\sigma(i-1)}=p \wedge \tau[p/\ivarx_{\sigma(i-1)}])\\
    \vee &\bigvee_{p=0}^{\delta-1} \bigg(\ivarx_{\sigma(i-1)}=\ivarx_{\sigma(i-2)}+p\wedge \tau[\ivarx_{\sigma(i-2)}+p/\ivarx_{\sigma(i-1)}]\bigg)\\
    \vee & \bigg( \ivarx_{\sigma(i-1)}\ge \gamma \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)}+\delta \wedge \ivarx_{\sigma(i-1)}\le \alpha(\vec\ivary)-1\bigg)\\
    \vee & \bigvee_{p=\alpha_1}^{\alpha_2+1} \bigg( \ivarx_{\sigma(i-1)}\ge \gamma \wedge \ivarx_{\sigma(i-1)} \ge \ivarx_{\sigma(i-2)}+\delta \wedge \ivarx_{\sigma(i-1)} = p \wedge \tau[p/\ivarx_{\sigma(i-1)}]\bigg),
\end{aligned}\end{equation}
where all exponential occurrences of $\ivarx_{\sigma(i-1)}$ are eliminated. Note that in the last line of (\ref{opt-example}), $\ivarx_{\sigma(i-1)}$ is  enumerated from $\alpha_1$ to $\alpha_2+1$, while in (\ref{lemma-example}) only two term $\alpha(\vec \ivary)$ and $\alpha(\vec \ivary)+1$ are tested for $\ivarx_{\sigma(i)}$.

### 2022-04-01 Acta Informatica2022/7_impl_exp.tex
.
before modify sec 4

\subsection{Implementation}
We implement the decision procedure in Wolfram Mathematica, called the {\paexp}-solver, which is able to solve the satisfiability of {\paexp} formulas.

The {\paexp}-solver takes a quantifier-free {\paexp} formula as input. Moreover, it allows specifying a upper bound $10^u$ to uniformly bound the values of variables. Given a upper bound $10^u$, the problem becomes to decide whether there is an assignment in which all variable values are bounded by $10^u$ satisfying the given formula.
%
Outputs of the {\paexp}-solver are either ``SAT'', ``UNSAT'', ``B-UNSAT'', or ``TIMEOUT'', standing for the given formula satisfiable, unsatisfiable, unsatisfiable up to $10^u$, or timeout if the solver does not terminate within the time limit. If the output is ``SAT'', then a model (namely, an assignment) is returned.

@@

To evaluate the performance of the {\paexp}-solver, we design two benchmark suites, ARITHMETIC and STRINGHASH\footnote{The benchmarks are available at https://github.com/EcstasyH/PAexp-Solver/tree/main/Benchmark}.
%\begin{itemize}
%\item

\paragraph{The ARITHMETIC benchmark suite} 
This suite comprises three groups of randomly generated {\paexp} formulas. Each group is characterized by four parameters ($EV$, $LV$, $EI$, $LI$), where $EV, LV$ represent the number of variables with exponential occurrences and with only linear occurrences respectively, and $EI, LI$ represent the number of inequalities with exponential terms and with only linear terms respectively. 
We consider three parameter classes, $(2, 3, 3, 4)$, $(3, 4, 4, 5)$, and $(4, 5, 5, 6)$. 
Each group of the benchmark suite consists of 200 randomly generated problem instances. The coefficients of exponential terms are randomly selected from the interval $[-10^2, 10^2]$ and the other coefficients/constants are randomly selected from $[-10^5, 10^5]$. The two intervals are chosen with the intention that the coefficients of exponential terms are smaller so that they do not always dominate the left-hand side of the inequalities. Moreover, aiming at a better coverage of the syntactical ingredients of {\paexp}, we randomly choose some problem instances and replace the $\le$ symbol of their first inequalities by $=$. The constant upper bound for the values of variables is set to be $10^{20}$, as the largest 64-bit integer is less than $10^{20}$. We also create an SMTLib2 file for each problem instance, to facilitate the comparison with the state-of-the-art of SMT solvers CVC4 and Z3. Because neither CVC4 nor Z3 supports the exponential functions directly, in the SMTLib2 files, we encode $10^x$ as a recursive function $f(x)$ defined by: $f(0) = 1$ and $f(n+1) = 10*f(n)$.

@@

The string constraints in the STRINGHASH benchmark suite are of the form 
%
$\svarx \in \Aut \wedge \left( \parseInt(\svarx) \bmod m \right) \bmod m^\prime = 0   \notag \wedge \len(\svarx) < 100,$ 
%
where $\Aut$ is an FA, $m, m^\prime \ge 2.$ 
The two groups of string constraints are characterized by flat and non-flat regular constraints respectively.
The flat group comprises 300 problem instances, where the flat languages are of the form $12345w^+_1 w^+_2$,  $12345 w^+_1  w^+_2 6789$, or $w^+_1 w^+_2 6789$, with $w_1,w_2 \in \Sigma^+_{\sf num}$, where $12345$ and $6789$ are the text to be protected, and $w^+_1 w^+_2$ is the pattern for nonce string. The non-flat group comprises 300 problem instances, where the non-flat languages are of the form $12345\Sigma_{\sf num}^*$, $12345\Sigma_{\sf num}^* 6789$, or $\Sigma_{\sf num}^* 6789$. Moreover, the number $m$ is a randomly chosen prime number in the interval $[10^2, 10^5]$ and $1 \le m' < m$ ($m'$ is not necessarily a prime number). We generate the SMTLib2 files for these string constraints, as inputs to the string constraint solvers. On the other hand, for the {\paexp}-solver, we do the following:   
\begin{itemize}
\item For flat instances, we generate {\paexp} formulas corresponding to the string constraints, as inputs to the {\paexp}-solver.
\item For non-flat instances, we use flat languages $a^* (b_{1} \ldots b_{k})$ to under-approximate 
$\Sigma_{\sf num}^*$,  where $a, b_1,\ldots, b_k \in \Sigma_{\sf num}$. 
%
%the first part is a single digit and the second part is a string length $l$, which can be seen as a FL with 2 cycles and force the second cycle do not repeat. 
%
We iterate the following procedure until a model is found or timeout: Initially, set $k=1$ and iterate by assigning $0, \ldots, 9$ to $a$. For each assignment, we encode the resulting string constraint into an {\paexp} formula with only one exponential variable. If the resulting {\paexp} formula is unsatisfiable, then we increase $k$ by 1 and repeat this process. 
\end{itemize}
We would like to remark that the flattening strategy for non-flat regular constraints here is a strict generalization of that in \cite{Parosh:20:PLDI}: Patterns of the form $0^*(b_1...b_k)$ were considered therein and {\pa} formulas are sufficient to encode such patterns. On the other hand, we consider patterns of the form e.g. $(a)^* (b_1 \ldots b_k)$ (where $a \in \Sigma_{\sf num}$ can be nonzero), which requires {\paexp} formulas to encode in general.

@@

We compare the {\paexp}-solver with the state-of-the-art SMT solvers on the generated benchmarks. Specifically, 
\begin{itemize}
\item for the ARITHMETIC benchmark suite, we compare the {\paexp}-solver against CVC4 (version 1.8) and Z3 (version 4.8.10),
\item for the STRINGHASH benchmark suite, we compare the {\paexp}-solver against CVC4, Z3, and Trau\footnote{\url{https://github.com/guluchen/z3/tree/new_trau}}. 
\end{itemize}
All the experiments are run on a lap-top with the Intel i5 1.4GHz CPU and 8GB memory. We set the time limit as 60 seconds per problem instance.

@@

For the ARITHMETIC benchmark suite, the $\paexp$-solver solves around $20\%$-$60\%$ more instances than Z3, and $30\%$-$100\%$ more instances than CVC4. Moreover, the gap becomes bigger as the the sizes of the formulas increase, which demonstrates that the $\paexp$-solver is more efficient in solving formulas of greater sizes. The average time of the $\paexp$-solver is comparable with Z3 and CVC4. The $\paexp$-solver reports ``B-UNSAT'' for 47 instances of the $(2, 3, 3, 4)$-group, while it does not report ``B-UNSAT'' (except one) for the other two groups. If more time is allowed, the $\paexp$-solver is able to report ``B-UNSAT'' for the ``TIMEOUT'' instances.

For the STRINGHASH benchmark suite, in overall, the $\paexp$-solver solves significantly more instances, especially those satisfiable instances, than Z3, CVC4, and Trau. For instance, for flat regular constraints, the $\paexp$-solver solves almost all 300 problem instances, except 3 of them\footnote{These three instances can actually be solved in 70 seconds.}, while  Z3, CVC4, Trau solve only 34, 89, 187 instances respectively.  Trau gets wrong answers for some problem instances, e.g. it reports ``UNSAT'' for some satisfiable instances. 
From the results, we can see that the $\paexp$-solver achieves a good tradeoff between precision and efficiency, although it is slower than the other solvers. 
%(More detailed experiment results of STRINGHASH can be found in Table~\ref{table:string}.)

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% Merge bound and unbound 
%\vspace*{-3mm}
\setlength\tabcolsep{3pt}

\begin{table*}[ht]
    \caption{Experimental Results. The left table summarizes the results for both experiments, and more detailded results of STRINGHASH benchmark suite are presented in the right table. O: Output, S:SAT, U: UNSAT, B: Bounded UNSAT, F: Fail, 

@@

        \multirow{2}{*}{Group }  & \multirow{2}{*}{O} & \multicolumn{2}{c}{Z3} & \multicolumn{2}{c}{CVC4} &  \multicolumn{2}{c}{Trau} & \multicolumn{2}{c }{$\paexp$} \\ 

@@

        \multirow{2}{*}{Group }  & \multirow{2}{*}{O} & \multicolumn{2}{c}{Z3} & \multicolumn{2}{c}{CVC4} &  \multicolumn{2}{c}{Trau} & \multicolumn{2}{c}{$\paexp$} \\

@@

>>>>

\subsection{Implementation}
We implement the decision procedure in Wolfram Mathematica, called EXP-solver, which is able to solve the satisfiability of {\paexp} formulas.

EXP-solver takes a quantifier-free {\paexp} formula as input. Moreover, it allows specifying a upper bound $10^u$ to uniformly bound the values of variables. Given a upper bound $10^u$, the problem becomes to decide whether there is an assignment in which all variable values are bounded by $10^u$ satisfying the given formula.
%
Outputs of EXP-solver are either ``SAT'', ``UNSAT'', ``B-UNSAT'', or ``TIMEOUT'', standing for the given formula satisfiable, unsatisfiable, unsatisfiable up to $10^u$, or timeout if the solver does not terminate within the time limit. If the output is ``SAT'', then a model (namely, an assignment) is returned.

@@

To evaluate the performance of EXP-solver, we design two benchmark suites, ARITHMETIC and STRINGHASH\footnote{The benchmarks are available at https://github.com/EcstasyH/PAexp-Solver/tree/main/Benchmark}.
%\begin{itemize}
%\item

\paragraph{The ARITHMETIC benchmark suite} 
This suite comprises three groups of randomly generated {\paexp} formulas. Each group is characterized by four parameters ($EV$, $LV$, $EI$, $LI$), where $EV, LV$ represent the number of variables with exponential occurrences and with only linear occurrences respectively, and $EI, LI$ represent the number of inequalities with exponential terms and with only linear terms respectively. 
We consider three parameter classes, $(2, 3, 3, 4)$, $(3, 4, 4, 5)$, and $(4, 5, 5, 6)$. 
Each group of the benchmark suite consists of 200 randomly generated problem instances. The coefficients of exponential terms are randomly selected from the interval $[-10^2, 10^2]$ and the other coefficients/constants are randomly selected from $[-10^5, 10^5]$. The two intervals are chosen with the intention that the coefficients of exponential terms are smaller so that they do not always dominate the left-hand side of the inequalities. Moreover, aiming at a better coverage of the syntactical ingredients of {\paexp}, we randomly choose some problem instances and replace the $\le$ symbol of their first inequalities by $=$. The constant upper bound for the values of variables is set to be $10^{20}$, as the largest 64-bit integer is less than $10^{20}$. We also create an SMTLib2 file for each problem instance, to facilitate the comparison with the state-of-the-art of SMT solvers CVC4 and Z3. Because neither CVC4 nor Z3 supports the exponential functions directly, in the SMTLib2 files, we encode $10^x$ as a recursive function $f(x)$ defined by: $f(0) = 1$ and $f(n+1) = 10\cdot f(n)$.

@@

The string constraints in the STRINGHASH benchmark suite are of the form 
%
$\svarx \in \Aut \wedge \left( \parseInt(\svarx) \bmod m \right) \bmod m^\prime = 0   \notag \wedge \len(\svarx) < 100,$ 
%
where $\Aut$ is an FA, $m, m^\prime \ge 2.$ 
The two groups of string constraints are characterized by flat and non-flat regular constraints respectively.
The flat group comprises 300 problem instances, where the flat languages are of the form $12345w^+_1 w^+_2$,  $12345 w^+_1  w^+_2 6789$, or $w^+_1 w^+_2 6789$, with $w_1,w_2 \in \Sigma^+_{\sf num}$, where $12345$ and $6789$ are the text to be protected, and $w^+_1 w^+_2$ is the pattern for nonce string. The non-flat group comprises 300 problem instances, where the non-flat languages are of the form $12345\Sigma_{\sf num}^*$, $12345\Sigma_{\sf num}^* 6789$, or $\Sigma_{\sf num}^* 6789$. Moreover, the number $m$ is a randomly chosen prime number in the interval $[10^2, 10^5]$ and $1 \le m' < m$ ($m'$ is not necessarily a prime number). We generate the SMTLib2 files for these string constraints, as inputs to the string constraint solvers. On the other hand, for EXP-solver, we do the following:   
\begin{itemize}
\item For flat instances, we generate {\paexp} formulas corresponding to the string constraints, as inputs to EXP-solver.
\item For non-flat instances, we use flat languages $a^* (b_{1} \ldots b_{k})$ to under-approximate 
$\Sigma_{\sf num}^*$,  where $a, b_1,\ldots, b_k \in \Sigma_{\sf num}$. 
%
%the first part is a single digit and the second part is a string length $l$, which can be seen as a FL with 2 cycles and force the second cycle do not repeat. 
%
We iterate the following procedure until a model is found or timeout: Initially, set $k=1$ and iterate by assigning $0, \ldots, 9$ to $a$. For each assignment, we encode the resulting string constraint into an {\paexp} formula with only one exponential variable. If the resulting {\paexp} formula is unsatisfiable, then we increase $k$ by 1 and repeat this process. 
\end{itemize}
We would like to remark that the flattening strategy for non-flat regular constraints here is a strict generalization of that in \cite{Parosh:20:PLDI}: Patterns of the form $0^*(b_1...b_k)$ were considered therein and {\pa} formulas are sufficient to encode such patterns. On the other hand, we consider patterns of the form e.g. $(a)^* (b_1 \ldots b_k)$ (where $a \in \Sigma_{\sf num}$ can be nonzero), which requires {\paexp} formulas to encode in general.

@@

We compare EXP-solver with the state-of-the-art SMT solvers on the generated benchmarks. Specifically, 
\begin{itemize}
\item for the ARITHMETIC benchmark suite, we compare EXP-solver against CVC4 (version 1.8) and Z3 (version 4.8.10),
\item for the STRINGHASH benchmark suite, we compare EXP-solver against CVC4, Z3, and Trau\footnote{\url{https://github.com/guluchen/z3/tree/new_trau}}. 
\end{itemize}
All the experiments are run on a lap-top with the Intel i5 1.4GHz CPU and 8GB memory. We set the time limit as 60 seconds per problem instance.

@@

\setlength\tabcolsep{3pt}

\begin{table*}[t]
    \caption{Experimental Results, left: results for both experiments, right: more detailded results of STRINGHASH benchmark suite. O: Output, S:SAT, U: UNSAT, B: Bounded UNSAT, F: Fail, 

@@

        \multirow{2}{*}{Group }  & \multirow{2}{*}{O} & \multicolumn{2}{c}{Z3} & \multicolumn{2}{c}{CVC4} &  \multicolumn{2}{c}{Trau} & \multicolumn{2}{c }{EXP-solver} \\ 

@@

        \multirow{2}{*}{Group }  & \multirow{2}{*}{O} & \multicolumn{2}{c}{Z3} & \multicolumn{2}{c}{CVC4} &  \multicolumn{2}{c}{Trau} & \multicolumn{2}{c}{EXP-solver} \\

@@

For the ARITHMETIC benchmark suite, EXP-solver solves around $20\%$-$60\%$ more instances than Z3, and $30\%$-$100\%$ more instances than CVC4. Moreover, the gap becomes bigger as the the sizes of the formulas increase, which demonstrates that EXP-solver is more efficient in solving formulas of greater sizes. The average time of EXP-solver is comparable with Z3 and CVC4. EXP-solver reports ``B-UNSAT'' for 47 instances of the $(2, 3, 3, 4)$-group, while it does not report ``B-UNSAT'' (except one) for the other two groups. If more time is allowed, EXP-solver is able to report ``B-UNSAT'' for the ``TIMEOUT'' instances.

For the STRINGHASH benchmark suite, in overall, EXP-solver solves significantly more instances, especially those satisfiable instances, than Z3, CVC4, and Trau. For instance, for flat regular constraints, EXP-solver solves almost all 300 problem instances, except 3 of them\footnote{These three instances can actually be solved in 70 seconds.}, while  Z3, CVC4, Trau solve only 34, 89, 187 instances respectively.  Trau gets wrong answers for some problem instances, e.g. it reports ``UNSAT'' for some satisfiable instances. 
From the results, we can see that EXP-solver achieves a good tradeoff between precision and efficiency, although it is slower than the other solvers. 
%(More detailed experiment results of STRINGHASH can be found in Table~\ref{table:string}.)

% Please add the following required packages to your document preamble:
% \usepackage{multirow}
% Merge bound and unbound 
%\vspace*{-3mm}
